---
title: "Intro_to_R_for_biologists"
author: "Jian"
date: "17/07/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load library
```{r}
library(tidyverse)
```

Read in breast cancer RNA-seq data
```{r}
counts = read_csv("GSE60450_GeneLevel_Normalized(CPM.and.TMM)_data.csv")
sampleInfo = read_csv("GSE60450_filtered_metadata.csv")
```

view what is stored in variables
```{r}
sampleInfo
counts
```

dimension of variables-> rows by columns
```{r}
dim(sampleInfo)
dim(counts)
```

view the first 6 lines by default or specify more lines through Arg
```{r}
head(sampleInfo)
?head # check Arg n
head(sampleInfo, 8)
```

view the last 6 lines
```{r}
tail(sampleInfo)
```

view the whole variable
```{r}
View(sampleInfo)# or just click on the variable in the Environment pane
View(counts)
```

view column vectors
```{r}
sampleInfo$X1
sampleInfo$immunophenotype
```

view values from a to b [a:b] in a column vector
```{r}
sampleInfo$X1[1:3]
sampleInfo$X1[2:4]
sampleInfo$immunophenotype[1:3]
```

view the structure of the data
```{r}
str(sampleInfo)
str(counts)
```

summary of data: length of string vectors refers to num of coordinates, whereas for numerical vectors: min, max, 1st quartile, 2nd quartile(median), 3rd quartile
```{r}
summary(counts)
summary(sampleInfo)
```

###  Excercises 1-4
```{r}
# 1.
?head
head(sampleInfo, n = 8)
# 2.
subsetCounts = head(counts, n = 20)
# 3.
subsetCounts$GSM1480291
# 4.
mean(subsetCounts$GSM1480291)
```

Formatting the data
```{r}
seqData = pivot_longer(counts, col = starts_with('GSM'), names_to = 'Sample', values_to = 'Count')

# or
seqData = pivot_longer(counts, col = GSM1480291:GSM1480302, names_to = 'Sample', values_to = 'Count')

# or
seqData = pivot_longer(counts, col = -c('X1', 'gene_symbol'), names_to = 'Sample', values_to = 'Count')

allInfo = full_join(seqData, sampleInfo, by = c('Sample' = 'X1'))
```

Plot data
```{r}
ggplot(allInfo, mapping = aes(x = Sample, y = log2(Count + 1), colour = Sample)) + 
  geom_violin()
```

```{r}
ggplot(allInfo, mapping = aes(x = Sample, y = log2(Count + 1), fill = Sample)) +
  geom_violin()
```

```{r}
ggplot(allInfo, mapping = aes(x = Sample, y = log2(Count + 1), 
                              fill = characteristics)) + geom_boxplot()
```

```{r}
ggplot(allInfo, mapping = aes(x = Sample, y = log2(Count + 1), 
                              fill = `developmental stage`)) + geom_boxplot()
```

Shorten Category names
```{r}
allInfo <-  mutate(allInfo, Group = case_when(
  str_detect(characteristics, 'basal.*virgin') ~ 'bvirg',
  str_detect(characteristics, 'basal.*preg') ~ 'bpreg',
  str_detect(characteristics, 'basal.*lact') ~ 'blact',
  str_detect(characteristics, 'luminal.*virgin') ~ 'lvirg',
  str_detect(characteristics, 'luminal.*preg') ~ 'lpreg',
  str_detect(characteristics, 'luminal.*lact') ~ 'llact',
))
```

Select 8 genes with the highest counts summed across all samples
```{r}
myGenes <-  allInfo %>%
  group_by(gene_symbol) %>%
  summarise(Total_count = sum(Count)) %>%  # remove repeated values
  arrange(desc(Total_count)) %>% # Arrange data into descending order
  head(n = 8) %>%
  pull(gene_symbol) # Pull out a single variable
```

Filter data
```{r}
myGenesCounts = filter(allInfo, gene_symbol %in% myGenes)
```

Create plot for each of the 8 genes
```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), fill = Group)) +
  geom_boxplot() +
  facet_wrap(~ gene_symbol)
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1),fill = Group)) +
  geom_point() +
  facet_wrap(~ gene_symbol)
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol)
```

customise plots
1. colours
```{r}
myColours = c('turquoise', 'plum', 'tomato', 'violet', 'steelblue', 'chocolate')

ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  scale_colour_manual(values = myColours)
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  scale_colour_brewer(palette = 'Dark2')
```

2. axis
```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'Count', 
       title = 'Mammary gland RNA-seq data')
```

3. theme
```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', 
       title = 'Mammary gland RNA-seq data') +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', 
       title = 'Mammary gland RNA-seq data') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', 
       title = 'Mammary gland RNA-seq data') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

```{r}
ggplot(data = myGenesCounts, 
       mapping = aes(x = Group, y = log2(Count + 1), colour = Group)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', 
       title = 'Mammary gland RNA-seq data') +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

order and categories(levels)
```{r}
groupOrder = c('bvirg', 'bpreg', 'blact', 'lvirg', 'lpreg', 'llact')

myGenesCounts = mutate(myGenesCounts, Group_f = factor(Group, levels = groupOrder))

ggplot(myGenesCounts, 
       mapping = aes(x = Group_f, y = log2(Count + 1), colour = Group_f)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', 
       title = 'Mammary gland RNA-seq data') +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

```

save the plot
```{r}
pdf('myplot.pdf')

ggplot(myGenesCounts, mapping = aes(x = Group_f, y = log2(Count + 1), colour = Group_f)) +
  geom_jitter() +
  facet_wrap(~ gene_symbol) +
  labs(x = 'Cell type and stage', y = 'log2(Count + 1)', title = 'Mammary gland RNA-seq data') +
  theme(axis.text.x = element_text(angle = 90)) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

dev.off()
```

The end
